Automatic Annotation of Tasks in Structured Code

### 2018-07-25 docs/ACM/paper.tex

OpenMP's runtime support allows the parallelization of irregular applications,
such as programs that traverse data-structures formed by a mesh of pointers.
In such programs, control constructs like {\tt if} statements make the
execution of some statements dependent on the program's input.
The runtime can capture such dependences, in contrast to static analysis tools.
As an example, Figure~\ref{fig:ex_book_filter} shows an application that
finds patterns in lines of a book\footnote{Whenever we show code, statements in
black font are part of the original program.
The annotations that we create automatically appear in grey.
The meaning of the pragmas used in the examples is explained in
Section~\ref{sec:sol}}.
The book is given as an array of pointers.
Each pointer leads to a string representing a potentially different line.
Our parallelization of this program consists in firing off a task
to process each line.
Figure~\ref{fig:ex_book_filter} has been annotated by
\Taskminer.
The OpenMP execution environment ensures the correct scheduling of the tasks
created at lines 9 and 11, by reassuring that the annotated dependencies
{\tt line} and {\tt pattern}
are respected at runtime.

@@

The two previous challenges are related to the performance of annotated
programs: if left unsolved, we shall have correct, albeit inefficient programs.
Challenges~\ref{ch:Regions}, and~\ref{ch:privatize}, in turn, are
related to correctness.
Challenge~\ref{ch:privatize} asks for the identification of variables that
must be replicated among threads.
This process of replication is called {\em privatization}.
As an example, variable \textsf{j} in Figure~\ref{fig:ex_privatize} must
be privatized.
In the absence of such action, that variable will be shared among all the
tasks created at line 5 of Figure~\ref{fig:ex_privatize}.
Because \textsf{j} is written to within these tasks, race conditions would
ensue.
Section~\ref{sub:variance} explains how we distinguish private from
shared variables.

@@

\paragraph{The annotation zoo.}
Our equipment of choice to exploit task parallelism in programs is OpenMP 4.0.
%This annotation system lets us indicate to compiler
%and runtime environment when and how tasks must be created.
We describe below which annotations we are using, and explain, informally, their
semantics.
Full overviews of the syntax and semantics of these annotations are publicly
available\footnote{For a quick overview, we refer the reader to the leaflet
``Summary of OpenMP 4.0 C/C++ Syntax", which is made available by the OpenMP
group.}; hence, we shall not dive into their details.
%
\begin{compactitem}
\item \textsf{parallel} (\textit{clauses}): forms a team of threads
which will execute the marked program region in parallel.
\item \textsf{single} (\textit{clauses}): specifies that a
program region must be executed by a single thread in the team.
\item \textsf{task} (\textit{clauses}): creates a new task.
\item \textsf{taskwait}: defines a synchronization point where
threads must wait for the completion of child tasks.
\end{compactitem}
%
Some pragmas are associated with lists of clauses.
There are several of these clauses available in OpenMP 4.0; however, we only use
the following:
\begin{compactitem}
\item \textsf{default(shared)}: indicates that variables are
shared by default among tasks.
\item \textsf{firstprivate($v$)}: indicates that $v$ must be replicated among
tasks, and initialized with its value at the point where the annotation is
executed.
\item \textsf{untied}: if a task has this modifier, then its associated code can
be executed by more than one thread; e.g., threads might alternate execution due
to preemption and load balancing.
\item \textsf{depend}($\mathit{in}/\mathit{out}/\mathit{inout}$):
determine if data is read ($\mathit{in}$), written ($\mathit{out}$) or both
within a task region; thus, effectively setting dependences among tasks.
\item \textsf{if}({\em condition}): this clause leads to the creation of an {\em undeferred}
task, which must be executed immediately.
We use it to enforce the cost model.
\item \textsf{final}({\em condition}) \textsf{mergeable}: this clause limits the creation of
tasks, and reuses the data environment of the father task.
We use it prune tasks considered too small, as discussed in Section~\ref{sub:ir}.
\end{compactitem}

@@

\begin{definition}[Symbolic Range Analysis]
\label{def:sra}
This is a form of abstract interpretation that associates
an integer variable $v$ with a {\em Symbolic Interval} $R(v) = [l, u]$, where
$l$ and $u$ are {\em Symbolic Expressions}.
A symbolic expression $E$ is defined by the grammar below, where $s$ is a
{\em program symbol}:
\renewcommand{\arraystretch}{0.9}
\[
\begin{array}{rcl}
E & ::= & z \ \ | \ \ s \ \ | E + E \ \ | \ \ E \times E \ \ | \ \  \mbox{min}(E, E) \ \ | \\
&  & \mbox{max}(E, E) \ \ | -\infty \ \ | \ \ +\infty
\end{array}
\]
\end{definition}

% How we do it.
The program symbols mentioned in Definition~\ref{def:sra} are names that cannot 
be reconstructed as functions of other names.
Examples include global variables, function arguments, and values returned by
external functions.
Symbolic range analysis is not a contribution of this paper: several
implementations of it exist.
We have adopted the one in the \dawn{} compiler~\cite{Mendonca17}, which, itself,
reuses work from Blume {\em et al.}~\cite{Blume94}.
The only extensions that we have added into \dawn 's implementation was
the ability to handle C-like \textsf{structs}.
Therefore, we shall not provide further details about this part of our work.
To understand the rest of this paper, it suffices to know that this implementation
is sufficiently solid to handle the entire C99 language, always terminates, and
runs in time linear on the size of the program's dependence graph.
Thus, in the worst case, it is quadratic on the number of program variables.
If a program contains an array access $\textsf{V}[E]$, and we have that
$R(E) = [l, u]$, then the memory region covered by this access is, at least,
$[\mathtt{\&}\textsf{V} + l, \mathtt{\&}\textsf{V} + u]$.
The example below clarifies this observation.

@@

\begin{definition}[Program Dependence Graph -- \cite{Ferrante87}]
\label{def:pdg}
Given a program $P$, its PDG contains one vertex for each statement $s \in P$.
There exists an edge from $s_1$ to $s_2$ if the latter depends on the former.
Statement $s_2$ is {\em data-dependent} on $s_1$ if it reads data that $s_1$
writes.
It is {\em control dependent} on $s_1$ if $s_1$ determines the outcome of a
branch, and depending on this outcome, we might execute $s_2$ or not.
\end{definition}

Task candidates are vanes in windmills.
{\em Windmills} are a family of graphs.
To the best of our knowledge, the term was coined by Rideau
{\em et al.}~\cite{Rideau08} to describe structural relations between register
copies.
Our windmills exist as subgraphs of the program's dependence graph.
We adopt a slightly more general definition than Rideaus'; however, the
metaphor that gave origin to the name: the shape of the graphs, is still
unmistakable:

@@

Figure~\ref{fig:ex_windmill} shows two program dependence graphs, highlighting
windmills and their vanes.
The structure of windmills naturally leads to task candidates.
Vanes correspond to program parts that are likely to be executed
several times, because they sprout from a loop -- the center of the windmill.
Two different executions of the same vane can run in parallel.
This is a property of the PDG, and one of the original motivations behind its
design.
These two different executions can be seen as two instances of the same
structure (the vane), coming out of the same windmill center.
A topological ordering of the dependence graph does not impose any order among
these two hypothetical replicas of the same structure.

@@

% Explain machinery: scope tree and simplification
\noindent
\textbf{The Scope Tree:}
To recover high-level information from the low-level IR, we have designed a
supporting data-structure henceforth named the {\em scope tree}.
The scope tree maps hammock regions into C constructions, such as
\textsf{while} and \textsf{if-then-else} blocks.
Each node of this graph represents a hammock region, augmented with
meta-information, such as the program part that it represents.
We keep track of these program parts via debugging information, which is
appended into the LLVM IR via the \textsf{-g} flag passed to the \textsf{clang}
compiler.
We have an edge from node $s_1$ to node $s_2$ if region $s_2$ is nested
within region $s_2$.
In the absence of control-flow optimizations such as loop unrolling or
dead-code elimination, each hammock region corresponds to some structured
code block (code region delimitable by braces).
Thus, we can find line numbers for each of the regions that we have marked
as tasks after the expansion seen in Section~\ref{sub:expansion}.

@@

There are more involved ways to bound the number of tasks.
We believe that the state-of-the-art in the field today is the work of
Iwasaki and Taura~\cite{Iwasaki16,Iwasaki16B}.
These authors propose different techniques to limit the creation of tasks,
be it through the replication of code, be it through the estimation of work,
given function inputs.
Pragmas for the conditional creation of tasks, such as the one seen in
lines 7 and 10 of Figure~\ref{fig:ex_cutoff}, lets us obtain much of the benefit
of code versioning, as proposed by Iwasaki and Taura.
However, we found work estimation of recursive functions a task too difficult
to accomplish in general.
Quoting~\cite[p.355]{Iwasaki16}: ``{\em there are tasks which essentially do not
have simple termination conditions (e.g., tasks traversing pointer-based
trees)}".
Thus, although simple, our recursion counters are general enough, handling even
such tasks that are hard to bound symbolically.
Nevertheless, we still would like to explore further ways to cut-off extremely
fine-grained tasks in the future.

@@

\begin{figure}[b!]
\begin{center}
\includegraphics[width=1\columnwidth]{images/TM_Performance}
\caption{Speedup comparisons -- automatic annotations compare favourably against
manual interventions, and lead to great performance gains.
The Y-axis shows the speedup of either manual intervention or the \Taskminer{}
(this paper) upon the original programs, in number of times.
Small numbers in boxes show speedup achieved by \Taskminer.}
\label{fig:TM_Performance}
\end{center}
\end{figure}

@@

The programs annotated by \Taskminer{} were faster than their sequential
counterparts in 13 of the 16 samples.
Most of the programs in these benchmarks were classic Divide and Conquer algorithms, such as \textsf{Strassen}'s matrix multiplication, \textsf{knapsack}
and \textsf{fft}.
In 6 of them, the automatically annotated programs were close and sometimes even 
slightly above the manually annotated versions.
In three examples where \Taskminer{} fell behind the original samples, e.g.,
\textsf{fib}, \textsf{floorplan} and \textsf{bellmandford} (\textsf{bell-ford}),
it produced code faster than the manually annotated competitor.
Visual inspection of \textsf{bellmanford} shows that the version parallelized
manually is dispatching a large number of small tasks.
Our cost model can prune some of these small tasks, but not enough to win
on the sequential version.
We emphasize that the human-annotated versions of our benchmarks have not
been tested originally in our architecture.
For instance, \textsf{BSC-Bots} was evaluated on an SGI Altix 4700
with 128 processors~\cite{Duran09}.
Thus, this difference in hardware might be accountable for the slowdown in
some of the manually annotated samples.
Nevertheless, this experiment demonstrates that {\em \Taskminer{} can
deliver non-negligible speedups comparable with sequential and manually
annotated programs}.

>>>>

OpenMP's runtime support allows the parallelization of irregular applications,
such as programs that traverse data-structures formed by a mesh of pointers.
In such programs, control constructs like {\tt if} statements make the
execution of some statements dependent on the program's input.
The runtime can capture such dependences, in contrast to static analysis tools.
As an example, Figure~\ref{fig:ex_book_filter} shows an application that
finds patterns in lines of a book\footnote{Whenever we show code, statements in
black font are part of the original program.
The annotations that we create automatically appear in grey.
The meaning of the pragmas used in the examples is explained in
Section~\ref{sec:sol}}.
The book is given as an array of pointers.
Each pointer leads to a string representing a potentially different line.
Our parallelization of this program consists in firing off a task
to process each line.
Figure~\ref{fig:ex_book_filter} has been annotated by
\Taskminer.
The OpenMP execution environment ensures the correct scheduling of the tasks
created at lines 10 and 12, by reassuring that the annotated dependencies
{\tt line} and {\tt pattern}
are respected at runtime.

@@

The two previous challenges are related to the performance of annotated
code: if left unsolved, we shall have correct, albeit inefficient programs.
Challenges~\ref{ch:Regions}, and~\ref{ch:privatize}, in turn, are
related to correctness.
Challenge~\ref{ch:privatize} asks for the identification of variables that
must be replicated among threads.
This process of replication is called {\em privatization}.
As an example, variable \textsf{j} in Figure~\ref{fig:ex_privatize} must
be privatized.
In the absence of such action, that variable will be shared among all the
tasks created at line 8 of Figure~\ref{fig:ex_privatize}.
Because \textsf{j} is written to within these tasks, race conditions would
ensue.
Section~\ref{sub:variance} explains how we distinguish private from
shared variables.

@@

\paragraph{The annotation zoo.}
Our equipment of choice to exploit task parallelism in programs is OpenMP 4.0.
%This annotation system lets us indicate to compiler
%and runtime environment when and how tasks must be created.
We describe below which annotations we are using, and explain, informally, their
semantics.
Full overviews of the syntax and semantics of these annotations are publicly
available\footnote{For a quick overview, we refer the reader to the leaflet
``Summary of OpenMP 4.0 C/C++ Syntax", which is made available by the OpenMP
ARB.}; hence, we shall not dive into their details.
%
\begin{compactitem}
\item \textsf{parallel} (\textit{clauses}): forms a team of threads
which will execute the marked program region in parallel.
\item \textsf{single} (\textit{clauses}): specifies that a
program region must be executed by a single thread in the team.
\item \textsf{task} (\textit{clauses}): creates a new task.
\item \textsf{taskwait}: defines a synchronization point where
threads must wait for the completion of child tasks.
\end{compactitem}
%
Some pragmas are associated with lists of clauses.
There are several of these clauses available in OpenMP 4.0; however, we only use
the following:
\begin{compactitem}
\item \textsf{default(shared)}: indicates that variables are
shared by default among tasks.
\item \textsf{firstprivate($v$)}: indicates that $v$ must be replicated among
tasks, and initialized with its value at the point where the annotation is
executed.
\item \textsf{untied}: if a task has this modifier, then its associated code can
be executed by more than one thread; e.g., threads might alternate execution due
to preemption and load balancing.
\item \textsf{depend}($\mathit{in}/\mathit{out}/\mathit{inout}$):
determine if data is read ($\mathit{in}$), written ($\mathit{out}$) or both
within a task region; thus, effectively setting dependences among tasks.
\item \textsf{if}({\em condition}): this clause leads to the creation of an {\em undeferred}
task, which must be executed immediately.
We use it to enforce the cost model.
\item \textsf{final}({\em condition}) \textsf{mergeable}: this clause limits the creation of
tasks, and reuses the data environment of the father task.
We use it prune tasks considered too small, as discussed in Section~\ref{sub:ir}.
\end{compactitem}

@@

\begin{definition}[Symbolic Range Analysis]
\label{def:sra}
This is a form of abstract interpretation that associates
an integer variable $v$ with a {\em Symbolic Interval} $R(v) = [l, u]$, where
$l$ and $u$ are {\em Symbolic Expressions}.
A symbolic expression $E$ is recursively defined as one of the terms:
$s, c, E + E, E \times E, \mbox{min}(E, E), \mbox{max}(E, E),$
$-\infty, +\infty$, where $s$ is a {\em program symbol}, and $c \in \mathbb{Z}$.
\end{definition}

The program symbols mentioned in Definition~\ref{def:sra} are names that cannot  be
reconstructed as functions of other names.
Examples include global variables, function arguments, and values returned by
external functions.
Symbolic range analysis is not a contribution of this paper: several
implementations of it exist.
We have adopted the one in the \dawn{} compiler~\cite{Mendonca17}, which, itself,
reuses an implementation by Nazar\'{e} {\em et al.}~\cite{Nazare14}.
The only extensions that we have added into \dawn 's implementation was
the ability to handle C-like \textsf{structs}.
Therefore, we shall not provide further details about this part of our work.
To understand the rest of this paper, it suffices to know that this implementation
is sufficiently solid to handle the entire C99 language, always terminates and runs in time 
linear on the size of the program's dependence graph.
Our implementation of Symbolic Range Analysis is interprocedural, to handle recursive 
functions, for instance.
Thus, in the worst case, it is quadratic on the number of program variables.
If a program contains an array access $\textsf{V}[E]$, and we have that
$R(E) = [l, u]$, then the memory region covered by this access is, at least,
$[\mathtt{\&}\textsf{V} + l, \mathtt{\&}\textsf{V} + u]$.
The example below clarifies this observation.

@@

\begin{definition}[Program Dependence Graph (PDG)~\cite{Ferrante87}]
\label{def:pdg}
The PDG of a program $P$ contains one vertex for each statement $s \in P$.
There exists an edge from $s_1$ to $s_2$ if the latter depends on the former.
Statement $s_2$ is {\em data-dependent} on $s_1$ if it reads data that $s_1$
writes.
It is {\em control dependent} on $s_1$ if $s_1$ controls a branch, whose result determines
if $s_2$ executes.
\end{definition}

Task candidates are vanes in windmills.
{\em Windmills} are a family of graphs.
The term was adopted by Rideau
{\em et al.}~\cite{Rideau08} to describe structural relations between register
copies.
Our windmills exist as subgraphs of the program's dependence graph.
We use a slightly more general definition than Rideaus'; however, the
metaphor that gave origin to the name: the shape of the graphs, is still
unmistakable:

@@

Figure~\ref{fig:ex_windmill} shows two program dependence graphs, highlighting
windmills and their vanes.
The structure of windmills naturally leads to task candidates.
Vanes correspond to program parts that are likely to be executed
several times, because they sprout from a loop -- the center of the windmill.
Two different executions of the same vane can run in parallel.
This is a property of the PDG, and one of the original motivations behind its
design.
These two different executions can be seen as two instances of the same
structure (the vane), coming out of the same windmill center.
A topological ordering of the dependence graph does not impose any order among
these two hypothetical replicas of the same structure.
Tasks can then be synchronized at the post-dominator of vanes marked to run in parallel
via the \textsf{taskwait} pragma.

@@

% Explain machinery: scope tree and simplification
\noindent
\textbf{The Scope Tree:}
To recover high-level information from the low-level IR, we have designed a
supporting data-structure henceforth named the {\em scope tree}.
The scope tree maps hammock regions into C constructions, such as
\textsf{while} and \textsf{if-then-else} blocks.
Each node of this graph represents a hammock region, augmented with
meta-information, such as the program part that it represents.
We keep track of these program parts via debugging information, which is
appended into the LLVM IR via the \textsf{-g} flag passed to the \textsf{clang}
compiler.
We have an edge from node $s_1$ to node $s_2$ if region $s_2$ is nested
within region $s_1$.
In the absence of control-flow optimizations such as loop unrolling or
dead-code elimination, each hammock region corresponds to some structured
code block (code region delimitable by braces).
Thus, we can find line numbers for each of the regions that we have marked
as tasks after the expansion seen in Section~\ref{sub:expansion}.

@@

There are more involved ways to bound the number of tasks.
We believe that the state-of-the-art in the field today is the work of
Iwasaki and Taura~\cite{Iwasaki16,Iwasaki16B}.
These authors propose different techniques to limit the creation of tasks,
be it through the replication of code, be it through the estimation of work,
given function inputs.
Pragmas for the conditional creation of tasks, such as the one seen in
lines 8 and 11 of Figure~\ref{fig:ex_cutoff}, lets us obtain much of the benefit
of code versioning, as proposed by Iwasaki and Taura.
However, we found work estimation of recursive functions a task too difficult
to accomplish in general.
Quoting~\cite[p.355]{Iwasaki16}: ``{\em there are tasks which essentially do not
have simple termination conditions (e.g., tasks traversing pointer-based
trees)}".
Thus, although simple, our recursion counters are general enough, handling even
such tasks that are hard to bound symbolically.
Nevertheless, we still would like to explore further ways to cut-off extremely
fine-grained tasks in the future.

@@

\begin{figure}[b!]
\begin{center}
\includegraphics[width=1\columnwidth]{images/TM_Performance}
\caption{Speedup comparisons -- automatic annotations compare favourably against
manual interventions, and lead to great performance gains.
The Y-axis shows the speedup of either manual intervention or the \Taskminer{}
(this paper) compared to the original programs, in number of times.
Small numbers in boxes show speedup achieved by \Taskminer.
The higher this number, the better.}
\label{fig:TM_Performance}
\end{center}
\end{figure}

@@

The programs annotated by \Taskminer{} were faster than their sequential
counterparts in 13 of the 16 samples.
Most of the programs in these benchmarks were classic Divide and Conquer algorithms, such as \textsf{Strassen}'s matrix multiplication, \textsf{knapsack}
and \textsf{fft}.
In 6 cases, automatically annotated programs were close or slightly faster than the manually
annotated versions.
In three examples where \Taskminer{} fell behind the original samples, e.g.,
\textsf{fib}, \textsf{floorplan} and \textsf{bellmandford} (\textsf{bell-ford}),
it produced code faster than the manually annotated competitor.
The version of \textsf{bellmanford} produced by \Taskminer{} is slower than its sequential
baseline.
The slowdown is not due to the cost model.
This algorithm traverses a graph implemented as an array of arrays, and data dependences
cause serialization at runtime.
%Visual inspection of \textsf{bellmanford} shows that the version parallelized
%manually is dispatching a large number of small tasks.
%Our cost model can prune some of these small tasks, but not enough to win
%on the sequential version.
We emphasize that the human-annotated versions of our benchmarks have not
been tested originally in our architecture.
For instance, \textsf{BSC-Bots} was evaluated on an SGI Altix 4700
with 128 processors~\cite{Duran09}.
Thus, hardware differences might be accountable for the slowdown in
some of the manually annotated samples.
Nevertheless, this experiment demonstrates that {\em \Taskminer{} can
deliver non-negligible speedups comparable with sequential and manually
annotated programs}.
